Ownership types in theory and practice (in Rust)

### 2022-11-30 paper.tex
Finished stack-based variable ownership transfer

The heap-allocated \emph{String} type in Rust provides a good example that demonstrates ownership transfer of heap-based values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% Since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.

\paragraph{Stack-based values}

In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
Therefore, stack-based values in Rust are copied instead of moved.
% TODO: reformulate
Internally, primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.

% The following example illustrates that the ownership of stack-based values is not transferred.
% \begin{lstlisting}
% let x = 5;
% let y = x;
% y = y + 1;
% assert_eq!(x, 5)
% \end{lstlisting}

The following piece of code demonstrates how primitive stack-based values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are

@@

% The two main representations of strings in Rust provide a good example.
% Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable. \cite{rust-by-example}

>>>>

The heap-allocated \emph{String} type in Rust provides a good example that demonstrates ownership transfer of heap-based values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.

% TODO: explain what Rust traits are
\paragraph{Stack-based values}

In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
% TODO: reformulate
Internally, this works because primitive types are annotated with and implement the \emph{Copy} trait.

% The two main representations of strings in Rust provide a good example.
Besides the heap-allocated \emph{String} type, Rust also supports immutable string literals \emph{\&str}, which are stored on the stack.

The following piece of code, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.

@@
